\section{Language and Typing}

We present a simple imperative language. 
\[
\begin{array}{lcll}
\text{Peripheral Declarations: }p & \bnfdef & \text{id} \{\langle s_{0},s_{1},\ldots,s_{n}\rangle, <, D \}, p \ \vert \ \cdot 
\end{array}
\]
\[
\begin{array}{lcll}
\text{Declarations: }d & \bnfdef & \text{fn } f(x, p) : (\tau_1, \Vec{s_1}) \rightarrow (\tau_2, \Vec{s_2}) \  \{c; ret := e\}, d \ \vert \ \cdot
\end{array}
\]
\[
\begin{array}{lcll}
\text{Commands: }c & \bnfdef & \text{let } x = e & \text{(let)} \\ 
  & \bnfalt &\text{let } x = f(e, p) & \text{(function call)} \\ & \bnfalt & \text{atomic}\{p_1:\Vec{s_1}, \dots, p_n:\Vec{s_n} \}(c) & \text{(atomic)} \\ 
  & \bnfalt & \text{if } e \text{ then } c_1 \text{ else } c_2 & \text{(branch)} \\ & \bnfalt & \text{transition}(x, \text{st}) & \text{(transition)} \\ & \bnfalt & \text{downgrade}(x, \text{st}) & \text{(downgrade)} \\ & \bnfalt & \text{clock} & \text{(clock update)} \\ & \bnfalt & \text{match }p \text{ with } st_1 \rightarrow c_1 | \dots | st_n \rightarrow c_n & \text{(match)} \\ & \bnfalt & c_1; c_2 & \text{(command seq)}
\end{array}
\]
\[
\begin{array}{lcll}
\text{Expressions: }e & \bnfdef & x & \text{(variable)}  \\
    & \bnfalt & v & \text{(value)} \\
    & \bnfalt & e_1 \text{ binop } e_2 & \text{(binary operation)} 
\end{array}
\]
Our judgments use the following additional constructs. 
\[
\begin{array}{llcl}
\textit{Typing Context} & \tctx & \bnfdef & x \rightarrow \tau, \tctx \bnfalt \cdot
\\ 
\textit{Types} & \tau & \bnfdef & \text{int} \bnfalt \text{bool} \bnfalt \text{unit} \bnfalt \rho \bnfalt (\tau_1, \Vec{s_1}) \rightarrow (\tau_2, \Vec{s_2})  
\\
\textit{Downgrade policy} & D & \bnfdef & s' \hookrightarrow s, D \bnfalt \cdot
\\
\textit{Peripheral Context} & \pctx & \bnfdef &  p \rightarrow \Vec{s}, \pctx \bnfalt \cdot
\end{array}
\]

A downgrade policy $s' \hookrightarrow s$ means that a peripheral in state $s'$ can transition to state $s$ assuming that $s' > s$. 
We also assume the existence of the cost functions 
$\m{cost}(s_{i},s_{j})$ = transition cost between \(s_{i},s_{j}\) when \(s_{i} \leq s_{j}\) or $\m{cost}(s_j, s_i)$ when \(s_j \hookrightarrow s_i\) and $\m{cost}(s_{i})$ = cost of being in state \(s_{i}\). Additionally, we assume $\m{cost}(\pctx)$ which returns the total cost of each peripheral being in the maximum possible state. 
We assume that the state lattice must have a unique \(\top\) element that is
the least upper bound of all the elements. 

For atomic regions, we assume the existence of a recovery system that can bring peripherals back to the states specified at the beginning of an atomic region (like Karma for example).
 
\sai{new AARA system starts here}

We want to parameterize the costs by two things: first, costs are "parametric" in the peripheral state. We mean that rather than tracking the peripheral typestate concretely, we update the peripheral context to be $\pctx ::= p : [\alpha], \pctx \mid \cdot$. When analyzing the cost of a function, we keep the costs generic in terms of the state, but when we have a function call, then we use the state inference rules to infer the concrete state of the peripheral, and output a concrete cost for the function based on the inferred state.

We have the following grammar for defining cost expressions:
\[
\begin{array}{lcll}
     \text{Cost expression} & C  \bnfdef & C(f) & \text{(command cost)} \\ & \bnfalt & C(p : st, f) & \text{(state cost)} \\ & \bnfalt & C(p : (st, st'), f) & \text{(transition cost)} \\ & \bnfalt & C(p : (\alpha, st), f) & \text{(non-concrete transition cost)} \\
     & \bnfalt & C(p : \alpha, f) & \text{(non-concrete cost)} \\ & \bnfalt & C_1 + C_2 & \text{(plus)} \\ & \bnfalt & \m{max}(C_1, C_2) & \text{(max)} \\ & \bnfalt & C[\alpha \mapsto st] & \text{(concretization)}
\end{array}
\]
We treat cost expressions as symbolic undetermined expressions. $\m{cost}(\pctx) := \sum_{p \in \pctx} C(p : \alpha)$ aka the sum of all of the peripherals in their current state variables. We have the following rule for joining peripheral contexts:

\begin{mathpar}
\inferrule*[left=Top Level Join]
{
\pctx_1 = p, \pctx_1' \\ \pctx_2 = p, \pctx_2' \\ \Phi_1, \Phi_2 \vdash \pctx_1[p] \uplus \pctx_2[p] = \pctx'[p], \Phi' \\ \Phi_1, \Phi_2, \Phi' \vdash \pctx_1' \uplus \pctx_2' = \pctx'
}
{
\Phi_1, \Phi_2, \cdot \vdash \pctx_1 \uplus \pctx_2 = \pctx', \Phi'
}

\inferrule*[left=Stack Join] 
{ \pi = \m{prefix}(\pctx_1[p], \pctx_2[p]) \\ \pctx_1[p] = \alpha_1 :: \sigma_1 :: \pi \\ \pctx_2[p] = \alpha_2 :: \sigma_2 :: \pi \\ \alpha' \text{ fresh} \\ \Phi' = (\Phi_1 \land \Phi_2) \land  (\alpha' = \alpha_1 \lor \alpha_2 ) \\ \pctx'[p] = \alpha' :: \sigma_1 :: \sigma_2 :: \pi
} 
{ \Phi_1, \Phi_2 \vdash \pctx_1[p] \uplus \pctx_2[p] = \pctx'[p], \Phi' 
}

\inferrule*[left=Empty]
{  }
{
\cdot \uplus \cdot = \cdot
}

\inferrule*[left=One Side Empty]
{ }
{\pctx_1 \uplus \cdot = \pctx_1}

\inferrule*[left=Symmetry]
{ }
{\pctx_1 \uplus \pctx_2 = \pctx_2 \uplus \pctx_1}
\end{mathpar}

We want to modell hardware-software concurrency as well. Following the lead of Abacus, we define transient states, which are states that represent when our peripheral is performing actions via hardware-initiated behavior (like interrupts). We update our language constructs in the following ways. First, we define the predicates transient($st$) and stable($st$) to specify when a state is either transient or stable. In more concrete terms, a state is transient if there is at least one \textbf{outgoing} transition that is hardware-initiated. To represent the concurrent behavior, we use an async/await syntax. We use capturing semantics to represent the idea that the peripheral is now executing concurrently and is not accessible to software. When we initiate an async call, the peripheral must be in a transient state, which is only possible from a software initiated transition. In order to model that transition into a transient state, we update the state inference constraint such that the new state must be a subset of $\{st\} \cup \text{HWReach}(st)$. What this means is that it must be a subset of the set of all reachable transient states from the state that is transitioned to. HWReach($st$) for a stable state $st$ is the empty set. When we typecheck an async command, we first solve for the input states and make sure the peripheral is in a transient state, then we typecheck the body of the asynchronous function - which represents the commands that are executed while in the transient state. Finally, we enforce that once the function is finished, the peripheral must transition back to a stable state. We then use the await construct to block execution until the thread has completed. However, blocking indefinitely would prevent us from being able to assign cost bounds to programs, as they can take abritrarily long periods of time. To get around this, we require that an await statement also includes a timeout and a fallback codeblock that executes once the timeout triggers. We then take the max cost of either triggering the timeout or completing execution. Finally, because a peripheral can now be in either a set of transient states, or a set of stable states, we require that a peripheral cannot be in a set of mixed states, i.e. both transient and stable. This means that anytime we check the state of a peripheral, it must either be all transient or all stable. Finally, we introduce $C_\text{transient}(p : \alpha)$ as a new cost, whic represents $\m{max}_{s \in \text{HWReach(}st\text{)}}(C(p : s))$ where $st$ = $\m{solve}(\alpha)$. We extend our language's commands to include the following:
\[
\begin{array}{lcll}
\text{Commands: }c & \bnfdef & \dots \\ 
  & \bnfalt &\text{let } tid = \text{async}(p) \ \{c\} & \text{(async peripheral operation)} \\ & \bnfalt & \text{await } tid \text{ timeout } \{c_\text{fb}\} & \text{(await)} 
\end{array}
\]

\begin{mathpar}
    \inferrule*[left=T-AsyncFun]
    { \pctx(p) = \alpha_i \\ \Vec{s_i} = \text{solve}(\Phi|_{\alpha_i}) \\ \forall st \in \vec{s_i}, \text{transient}(st) \\ \tctx, p : [\alpha_i], \Phi|_p, \cdot, f \vdash c \rightarrow \tctx_a, \pctx_a, p : [\alpha_o], C_a, f_a \\ \vec{s_o} = \text{solve}(\Phi_a|_{\alpha_o}) \\ \forall st \in \vec{s_o},\ \text{stable}(st) \\ \pctx' = \pctx \setminus \pctx|_p \\ \Phi' = \Phi \setminus \Phi|_p \\ C' = C + C_\text{async spawn}
    }
    { \tctx, \pctx, \Phi, C, f \vdash \text{let } tid = \text{async}(p)\ \{c\} \rightarrow \tctx[tid : (p, \pctx_a, \Phi_a, C_a, f_a)], \pctx', \Phi', C', f
    }
    
    \inferrule*[left=T-Await-Timeout]
    {\tctx(tid) = (p, \pctx_a, \Phi_a, C_a, f_a) \\ p \notin \pctx \\  \tctx, \pctx, \Phi, C, f \vdash c_\text{fb} \rightarrow \tctx_\text{fb}, \pctx_\text{fb}, \Phi_\text{fb}, C_\text{fb}, f_\text{fb} \\ \Vec{s_\text{fb}} = \text{solve}(\Phi_\text{fb}|_p) \\ \forall st \in \vec{s_\text{fb}}, \text{stable}(st) \\ C_\text{timeout} = \text{timeout * } C_\text{transient}(p : \alpha_i) \\ C_\text{await timeout} = C_\text{timeout}(f_\text{fb}) + C_\text{fb}(f_\text{fb}) \\ C' = C + \max(C_a(f_a), C_\text{await timeout}) \\ \pctx_r = \pctx \cup \pctx_a \\ \pctx'', \Phi'' = \pctx_r \uplus \pctx_\text{fb}
    }
    { \tctx, \pctx, \Phi, C, f \vdash \text{await } tid \text{ timeout} \ \{c_\text{fb}\} \rightarrow \tctx, \pctx'', \Phi'', C', f \cup f_a \cup f_\text{fb}
    }

    \inferrule*[left=T-Transition-Transient]
    { \tctx, \pctx, \Phi \vdash x : \rho \\ st = \text{transient}(st) \\ \pctx(x) = \alpha_0 \\ \alpha_1 \text{ fresh} \\ \Vec{st}^+ = st \cup \text{HWReach}(st) \\ \Phi' = \Phi \land \alpha_0 \leq st \land \alpha_1 \subseteq \Vec{st}^+ \\ \pctx'[x] = \alpha_1 :: \pctx[x] \\ C' = C(p : (\alpha_0, st), f)
    }
    { \tctx, \pctx, \Phi, C, f \vdash \text{transition}(x, st) \rightarrow \tctx, \pctx', \Phi', C', f
    }
\end{mathpar}

We represent the machine state as a tuple $\Sigma ::= (M, T, C, S)$ where $M$ is the memory mapping variables to values, $T$ is the thread pool to spawn new threads, $C$ is the stack of frames executing, $S$ is the map of peripherals to their state, and $T$ maps handles to a tuple $(p, c)$ which is the peripheral $p$ and the commands being executed. We model concurrent execution in a preemptive fashion. Our execution is instead a stack of frames, which each represent either a computation or a hole waiting for another thread to finish. We can spawn threads using $\m{async}$ which enter the thread pool, and the scheduler can non-deterministically choose a thread from the pool to be pushed onto the execution stack. We always execute commands from the top of the execution stack. We assume that async peripheral operations are \textit{capturing} - meaning that the peripheral is no longer available to the main program until the thread ends. We can use this to model hardware concurrency and transient states, since while the peripheral is captured, it performs it's own operations that the main thread is oblivious to.

\[
\begin{array}{llcl}
\textit{Memory} & M & \bnfdef & x \mapsto v, M \bnfalt \cdot
\\ 
\textit{Thread Pool} & T & \bnfdef & tid \mapsto (p, c), T \bnfalt \cdot 
\\
\textit{Peripheral Map} & S & \bnfdef & p \mapsto s, P \bnfalt \cdot
\\
\textit{Execution Context} & C & \bnfdef &  f, C \bnfalt \cdot
\\
\textit{Frame} & f & \bnfdef & (tid : p, c) \bnfalt \cdot
\end{array}
\]

\begin{mathpar}
    \inferrule*[left=E-Seq]
    {
    c_1 \rightarrow c'_1
    }
    {
    (M, T, (tid: p, c_1; c_2) :: C, S) \rightarrow (M, T, (tid: p, c'_1; c_2) :: C, S)
    }

    \inferrule*[left=E-SeqDone]
    { }
    {
    (M, T, (tid: p, \text{skip}; c_2) :: C, S) \rightarrow (M, T, (tid: p, c_2) :: C, S)
    }
    
    \inferrule*[left=E-Async]
    {
    tid \notin \m{dom}(T)
    }
    {
    (M, T, (tid' : p, \text{let } tid = \text{async}(p) \ \{c_a\}) :: C, S) \rightarrow (M, T[tid \mapsto (p, c_a)], (tid' : p, \text{skip}) :: C, S)
    }
    
    \inferrule*[left=E-AsyncPreempt]
    { }
    {
    (M, T[tid \mapsto (p, c_a)], C, S) \rightarrow (M, T \setminus tid, (tid: p, c_a) :: C, S)
    }

    \inferrule*[left=E-AsyncDone]
    {
    C = (tid: p, v) :: C' \\ tid \neq \text{main}
    }
    {
    (M, T, C, S) \rightarrow (M, T[tid \mapsto \text{done}(p, v)], C', S)
    }

    \inferrule*[left=E-Await]
    {
    T(tid) = \text{done}(p, v) 
    }
    {
    (M, T, (tid' : p, \text{let } x = \text{await } tid) :: C, S) \rightarrow (M[x \mapsto v], T, (tid' : p, \text{skip}) :: C, S)
    }

\end{mathpar}
\newpage
Our typing judgment is of the form: 
$$ \begin{aligned}
    \fbox{$\tctx, \pctx, \Phi, C, f \vdash c \rightarrow \tctx', \pctx', \Phi', C', f'$}
\end{aligned} $$
\begin{mathpar}
\inferrule*[left=T-Seq]
{
\tctx,\pctx,\Phi,f,C \vdash c_1 \rightarrow \tctx',\pctx',\Phi',f_1, C' \\
\tctx',\pctx',\Phi',C', f' \vdash c_2 \rightarrow \tctx'',\pctx'',\Phi'', C'', f'' \\ C_2 = C' + C'' + \m{cost}(\pctx)(f)
}
{
\tctx,\pctx,\Phi,f,C \vdash c_1\,;\,c_2 \rightarrow \tctx'',\pctx'',\Phi'', C_2, f_2
}

\inferrule*[left=T-If]
{
\tctx,\pctx,f \vdash e : \text{bool}, C_e(f) \\
\tctx,\pctx,\Phi,C, f \vdash c_1 \rightarrow \tctx_1,\pctx_1,\Phi_1, C_1, f_1 \\
\tctx,\pctx,\Phi, C, f \vdash c_2 \rightarrow \tctx_2, \pctx_2, \Phi_2, C_2, f_2 \\
\tctx^* = \tctx_1 \cup \tctx_2 \\
\pctx^*, \Phi^* = \pctx_1 \uplus \pctx_2 \\ C^* = \m{max}(C_1(f_1), C_2(f_2)) + C_e
}
{
\tctx,\pctx,\Phi, C, f \vdash
\text{if } e \text{ then } c_1 \text{ else } c_2
\rightarrow
\tctx^*, \pctx^*, \Phi^*, C^*, \{f_1, f_2\}
}

\inferrule*[left=T-Clock]
{
f' \text{ fresh}
}
{
\tctx,\pctx,\Phi, C, f \vdash \text{clock} \rightarrow \tctx,\pctx,\Phi, C_{\m{clock}}(f), f'
}

\inferrule*[left=T-Transition]
{
\tctx, \pctx, \Phi \vdash x : \rho \\
\pctx(x) = \alpha_0 \\ 
\alpha_1 \text{ fresh} \\
\Phi' = \Phi \land \alpha_0 \leq st \land \alpha_1 = st \\
\pctx'[x] = \alpha_1 :: \pctx[x] \\ C' = C(p : (\alpha_0, st), f)
}
{
\tctx,\pctx,\Phi, C, f \vdash \text{transition}(x, st) \rightarrow \tctx,\pctx',\Phi', C', f
}

\inferrule*[left=T-Downgrade]
{
\tctx, \pctx, \Phi \vdash x : \rho \\
s \hookrightarrow st \in x(D) \\
\pctx(x) = \alpha_0 \\ 
\alpha_1 \text{ fresh} \\
\Phi' = \Phi \land \alpha_0 = s \land \alpha_1 = st \\
\pctx'[x] = \alpha_1 :: \pctx[x] \\ C' = C(p : (\alpha_0, st),\, f)
}
{
\tctx, \pctx, \Phi, C, f \vdash \text{downgrade}(x, st) \rightarrow \tctx, \pctx', \Phi', C', f
}

\inferrule*[left=T-FunDecl]
{
\alpha_i \text{ fresh} \\
\pctx = p : [\alpha_i] \\
\tctx, x : \tau_1, \pctx, \cdot, f, \cdot \vdash c \rightarrow \pctx_f,\Phi_f, C, f_f \\
\pctx_f(p) = \alpha_o \\
\vec{s_1} = \m{solve}(\Phi_f|_{\alpha_i}) \\ \forall s \in \vec{s_1}, \text{stable}(s) \lor \forall s \in \vec{s_1}, \text{transient}(s) \\ 
\vec{s_2} = \m{solve}(\Phi_f|_{\alpha_o}) \\ \forall s \in \vec{s_2}, \text{stable}(s) \\
\tau_f = \tau_1 \rightarrow \tau_2 \\ \tctx, \pctx, f_f \vdash e: \tau_2, C_r(f_f) \\ C_f = C + C_r(f_f)
}
{
\tctx \vdash \text{fn } f(x, p) \ \{c; ret := e\} : \tau_f, C_f, \Phi_f|_{\alpha_i}, \vec{s_1}, \vec{s_2}, f_f
}

\inferrule*[left=T-Call]
{
\tctx(f) = \tau_1 \rightarrow \tau_2, C_f, \Phi_f|_{\alpha_i}, \vec{s_1}, \vec{s_2}, f' \\
\alpha = \pctx(p) \\
\vec{s} = \m{solve}(\Phi_f|_{\alpha_i} \land \Phi|_{\alpha}) \\  
  (\Vec{s} \subseteq \Vec{s_1} \text{ if } \forall st \in \Vec{s}, \text{ stable}(st)) \lor
  (\Vec{s} \cap \Vec{s_1} \neq \emptyset \text{ if }\forall st \in \Vec{s}, \text{ transient}(st)) \\ 
\tctx,\pctx, f \vdash e : \tau_1, C_a(f) \\
\alpha' \text{ fresh} \\
\pctx'[p] = \alpha' :: \pctx[p] \\
\Phi' = \Phi \land \alpha \subseteq \vec{s_1} \land \alpha' \subseteq \vec{s_2} \\ C' = C_a(f) + C_f[\Vec{s}]
}
{
\tctx,\pctx,\Phi,C, f \vdash \text{let } x = f(e, p) \rightarrow \tctx[x : \tau_2],\pctx',\Phi', C', f'
}

\end{mathpar}
\newpage
\sai{state inference rules start here}

The intuition for the state inference rules is that peripherals start in an unconstrained state. Function declarations give us constraints on the input and output states of a peripheral. However those constraints may be more general than the actual state of a peripheral when the function is invoked at a function call. So we want to delay the cost determination of a function until it is actually called. To get more accurate cost bounds, we want to parameterize the cost in terms of the possible peripheral states, and then infer the smallest possible set of peripheral states at the call site that still satisfy the function declaration, and concretize the cost there.

For atomic regions, we only care about satisfying the constraints on the possible peripheral states at the start of each region.

We have the following additional inference contexts:
\[
\begin{array}{lcll}
     \textit{Constraints} & \Phi & \bnfdef & \varphi, \Phi \bnfalt \cdot  \\
    \textit{State Constraints} & \varphi & \bnfdef & \alpha = \Vec{st} \bnfalt \alpha \leq st \bnfalt \alpha \hookrightarrow st \bnfalt \alpha \subseteq \Vec{st} \bnfalt \varphi_1 \wedge \varphi_2 \bnfalt \varphi_1 \vee \varphi_2 \\
    \textit{Peripheral State Context} & A & \bnfdef & p : [\alpha], A \bnfalt \cdot
\end{array}
\]

We define [$\alpha$] to be a stack of type variables, with a lookup $A(p)$ referring to the top of it's stack, $\alpha :: A(p)$ pushing $\alpha$ to the top of the stack and $A[p]$ to refer to the current stack of type variables for peripheral $p$. We also define $\m{prefix}(A_1, A_2)$ to return the common prefix of two contexts $A_1, A_2$ if one exists, or $\cdot$ if none exists.

We define the join between two peripheral state contexts, $A_1 \uplus A_2$ as follows.
\begin{mathpar}
\inferrule*[left=Top Level Join]
{
A_1 = p, A_1' \\ A_2 = p, A_2' \\ \Phi_1, \Phi_2 \vdash A_1[p] \uplus A_2[p] = A'[p], \Phi' \\ \Phi_1, \Phi_2, \Phi' \vdash A_1' \uplus A_2' = A'
}
{
\Phi_1, \Phi_2, \cdot \vdash A_1 \uplus A_2 = A', \Phi'
}

\sai{I am using $\sigma_1, \sigma_2$ to represent intermediate states that are between the prefix and the top of the stack}

\inferrule*[left=Stack Join] 
{ \pi = \m{prefix}(A_1[p], A_2[p]) \\ A_1[p] = \alpha_1 :: \sigma_1 :: \pi \\ A_2[p] = \alpha_2 :: \sigma_2 :: \pi \\ \alpha' \text{ fresh} \\ \Phi' = (\Phi_1 \land \Phi_2) \land \{ \alpha' = \alpha_1 \lor \alpha_2 \} \\ A'[p] = \alpha' :: \sigma_1 :: \sigma_2 :: \pi
} 
{ \Phi_1, \Phi_2 \vdash A_1[p] \uplus A_2[p] = A'[p], \Phi' 
}

\inferrule*[left=Empty]
{  }
{
\cdot \uplus \cdot = \cdot
}

\inferrule*[left=One Side Empty]
{ }
{A_1 \uplus \cdot = A_1}

\inferrule*[left=Symmetry]
{ }
{A_1 \uplus A_2 = A_2 \uplus A_1}
\end{mathpar}

We also define $\m{solve}(\varphi, \varepsilon)$ to a function that takes in a set of constraints $\varphi$ and the current partial solution $\varepsilon :: = \alpha \rightarrow \Vec{st}, \varepsilon$. It then returns $\varepsilon'$ which is the updated solution with the new constraints form solving $\varphi$. We have a top level function $\m{sol}(\Phi)$ which calls $\m{solve}$ on each $\varphi \in \Phi$. The solution satisfies the following constraints: \sai{i dont think these rules are exactly right but im trying to formalize my intuition}
\begin{mathpar}

\inferrule*[left=Solve-Empty]
{}
{\cdot \vdash \alpha \Downarrow \top}

\inferrule*[left=Solve-Eq]
{}
{\alpha = \Vec{s} \vdash \alpha \Downarrow \Vec{s}}

\inferrule*[left=Solve-Leq]
{}
{\alpha \leq st \vdash \alpha \Downarrow \{\, s \mid s \leq st \,\}}

\inferrule*[left=Solve-Subset]
{}
{\alpha \subseteq \Vec{s} \vdash \alpha \Downarrow \Vec{s}}



\inferrule*[left=Solve-And]
{
\varphi_1 \vdash \alpha \Downarrow S_1 \\
\varphi_2 \vdash \alpha \Downarrow S_2
}
{
\varphi_1 \wedge \varphi_2 \vdash \alpha \Downarrow S_1 \cap S_2
}

\inferrule*[left=Solve-Or]
{
\varphi_1 \vdash \alpha \Downarrow S_1 \\
\varphi_2 \vdash \alpha \Downarrow S_2
}
{
\varphi_1 \vee \varphi_2 \vdash \alpha \Downarrow S_1 \cup S_2
}


\inferrule*[left=Solve]
{
\varphi, \Phi' = \Phi \\ \varepsilon = \m{solve}(\varphi, \cdot) \\ \Phi', \varepsilon \vdash \m{sol}(\Phi)
}
{
\Phi, \cdot \vdash \m{sol}(\Phi) \Downarrow \epsilon
}

\end{mathpar}

Command state inference rules are of the form $$
\begin{aligned}
    \fbox{$\tctx, A, \Phi \vdash c \rightarrow A', \Phi'$}
\end{aligned}
$$


\begin{mathpar}
\inferrule*[left=T-Transition\(_\textit{inf}\)] 
{ \tctx, A, \Phi \vdash x : \rho \\ A(x) = \alpha_0 \\ \alpha_1 \text{ fresh} \\ \Phi' = \Phi \land \alpha_0 \leq st \land \alpha_1 = st \\ A'[x] = \alpha_1 :: A[x]
} 
{ 
\tctx, A,\Phi \vdash \text{transition}(x, st) : \text{unit} \rightarrow A', \Phi' 
}

\inferrule*[left=T-Downgrade\(_\textit{inf}\)]
{
\tctx, A, \Phi \vdash x : \rho \\ s \hookrightarrow st \in x(D) \\ A(x) = \alpha_0 \\ \alpha_1 \text{ fresh} \\ \Phi' = \Phi \land \alpha_0 = s \land \alpha_1 = st \\ A'[x] = \alpha_1 :: A[x]
}
{ 
\tctx, A, \Phi \vdash \text{downgrade}(x, st) : \text{unit} \rightarrow A', \Phi'
}

\inferrule*[left=T-If\(_\textit{inf}\)]
{
\tctx \vdash e : \text{bool} \\ \tctx, A, \Phi \vdash c_1 : \rightarrow A_1, \Phi_1\\ \tctx, A, \Phi \vdash c_2 \rightarrow A_2, \Phi_2 \\  A^*, \Phi^* = A_1 \uplus A_2
}
{
\tctx, A, \Phi \vdash \text{if } e \text{ then } c_1 \text{ else } c_2: \tau \rightarrow A^*, \Phi^*
}

\inferrule*[left=T-Seq\(_\textit{inf}\)]
{
\tctx, A, \Phi \vdash c_1 \rightarrow A', \Phi' \\ \tctx, A', \Phi' \vdash c_2 \rightarrow A'', \Phi''
}
{
\tctx, A, \Phi \vdash c_1; c_2 \rightarrow A'', \Phi''
}

\inferrule*[left=T-Call\(_\textit{inf}\)]
{
\tctx(f) = (\tau_1, \Vec{s_1}) \rightarrow (\tau_2, \Vec{s_2}) \\ \tctx \vdash e : \tau_1 \\ A(p) = \alpha \\ \alpha' \text{ fresh} \\ \Phi' = \Phi \land \alpha \subseteq \Vec{s_1} \land \alpha' \subseteq \Vec{s_2} \\ A'[p] = \alpha' :: A[p]
}
{
\tctx, A, \Phi \vdash \text{let } x = f(e, p) \rightarrow A', \Phi'
}

\inferrule*[left=T-Atomic\(_\textit{inf}\)]
{
\forall i \in [1, \dots, n], A(p_i) = \alpha_i \\ \varphi = \bigwedge_{i = 1}^n \alpha_i \subseteq \Vec{s_i} \\ \Phi^* = \Phi \wedge \varphi \\ \tctx, A, \Phi^* \vdash c \rightarrow A', \Phi'
}
{
\tctx, A, \Phi \vdash \text{atomic}\{p_1: \Vec{s_1}, \dots, p_n: \Vec{s_n} \}(c) \rightarrow A', \Phi'
}

\inferrule*[left=T-Fun\(_\textit{inf}\)] 
{ \alpha_{\text{in}} \text{ fresh} \\ A = p : [\alpha_{\text{in}}] \\ \tctx, A, \cdot \vdash c \rightarrow A_f, \Phi_f \\ A_f(p) = \alpha_{\text{out}} \\ \Vec{s}_o = \m{solve}(\Phi_f, \alpha_{\text{out}}) \\ \Vec{s}_i = \m{solve}(\Phi_f, \alpha_{\text{in}}) \\ \tau_f = (\tau_1, \Vec{s}_i) \rightarrow (\tau_2, \Vec{s}_o)
} 
{ \tctx \vdash \text{fn } f(x, p) \ \{c; ret := e\} : \tau_f }

\end{mathpar}

\sai{state inference rules end here}
\newpage
\sai{Typestate system starts here}

Command typing rules are of the form $$
\begin{aligned}
    \fbox{$\tctx, \pctx \vdash c: \tau \Rightarrow \tctx', \pctx'$}
\end{aligned}
$$

Expression typing rules are of the form $$
\begin{aligned}
    \fbox{$\tctx, \pctx \vdash e: \tau $}
\end{aligned}
$$

\begin{mathpar}

\inferrule*[left=T-Var]
{ \tctx(x) = \tau }
{ \tctx, \pctx \vdash x : \tau }


\inferrule*[left=T-BinOp]
{
\tctx,\pctx \vdash e_1 : \tau_1 \\ \tctx,\pctx \vdash e_2 : \tau_2 \\ \text{binop} : \tau_1 \times \tau_2 \to \tau
}
{
\tctx,\pctx \vdash e_1 \ \text{binop}\ e_2 : \tau
}

\inferrule*[left=T-Let]
{ \tctx,\pctx \vdash e : \tau }
{ 
\tctx,\pctx \vdash \text{let } x = e : \text{unit} \Rightarrow \tctx[x : \tau] , \pctx
}

\inferrule*[left=T-Call]
{
\tctx(f) = (\tau_1, \Vec{s_1}) \rightarrow (\tau_2, \Vec{s_2}) \\ \tctx,\pctx \vdash e : \tau_1 \\ \pctx(p) = \Vec{s} \\ \Vec{s} \subseteq s_1
}
{
\tctx,\pctx \vdash \text{let } x = f(e, p) : \text{unit} \Rightarrow \tctx[x : \tau_2], \pctx[p: s_2]
}

\inferrule*[left=T-Transition]
{
\tctx,\pctx \vdash x : \rho \\ \pctx(x) = \Vec{s} \\ \forall s_i\in \Vec{s}, s_i \leq st
}
{ 
\tctx,\pctx \vdash \text{transition}(x, st) : \text{unit} \Rightarrow \tctx, \pctx[x : st]
}

\inferrule*[left=T-Downgrade]
{
\tctx,\pctx \vdash x : \rho \\ \pctx(x) = \Vec{s} \\ \forall s_i\in \Vec{s}, s_i \hookrightarrow st \in p(D)
}
{ 
\tctx,\pctx \vdash \text{downgrade}(x, st) : \text{unit} \Rightarrow \tctx, \pctx[x : st]
}

\inferrule*[left=T-If]
{
\tctx,\pctx \vdash e : \text{bool} \\ \tctx,\pctx \vdash c_1 : \tau \Rightarrow \tctx_1, \pctx_1 \\ \tctx,\pctx \vdash c_2 : \tau \Rightarrow \tctx_2, \pctx_2 \\ \tctx^* = \tctx_1 \cup \tctx_2 \\ \pctx^* = \pctx_1 \cup \pctx_2 
}
{
\tctx, \pctx \vdash \text{if } e \text{ then } c_1 \text{ else } c_2 \Rightarrow \tctx^*, \pctx^*
}

\inferrule*[left=T-Seq]
{
\tctx,\pctx \vdash c_1 : \tau_1 \Rightarrow \tctx', \pctx' \\ \tctx',\pctx' \vdash c_2 : \tau_2 \Rightarrow \tctx'', \pctx''
}
{
\tctx,\pctx \vdash c_1 ; c_2 : \tau_2 \Rightarrow \tctx'', \pctx''
}

\inferrule*[left=T-Atomic]
{
\forall i, \pctx(p_i) \subseteq \Vec{s_i} \\ \tctx,\pctx \vdash c : \tau \Rightarrow \tctx', \pctx' 
}
{ 
\tctx,\pctx,S \vdash \text{atomic}\{p_1: \Vec{s_1}, \dots, p_n: \Vec{s_n} \}(c) : \tau \Rightarrow \tctx', \pctx'
}

\inferrule*[left=T-FunDecl]
{
\tctx[x : \tau_1], \pctx[p: \Vec{s_1}] \vdash c: \tau_2 \Rightarrow \tctx', \pctx' \\ \pctx'(p) = \Vec{s_2} \\ \tctx', \pctx' \vdash e : \tau_2
}
{ \tctx,\pctx \vdash \text{fn } f(x, p) : (\tau_1, \Vec{s_1}) \rightarrow (\tau_2, \Vec{s_2}) \  \{c; ret := e\} \Rightarrow \tctx[f \mapsto (\tau_1, \Vec{s_1}) \rightarrow (\tau_2, \Vec{s_2})], \pctx
}

\end{mathpar}

\sai{Typestate system ends here}