\documentclass[sigconf,nonacm,screen,review]{acmart}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathpartir}
\usepackage{listings}
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}


\include{macros}
\begin{document}


\title{Automatic Energy Analysis Using Types}
\author{Sai Divvela}
\setcctype{none}
\settopmatter{printacmref=false, printccs=false}
\maketitle


\section{Problem and Motivation}
As IoT devices become increasingly prevalent, embedded systems operating in resource-constrained environments must carefully account for their energy requirements. These requirements are typically expressed as the \textit{worst-case energy consumption} (WCEC), the maximum possible energy consumed across a single execution, and are obtained either through measurement-based profiling or static analysis techniques that reason over all possible program executions.

Both approaches face significant challenges when peripheral devices such as sensors, radios, and actuators are involved. Peripherals maintain internal state that evolves based on the operations they perform, making them difficult to reason about. Measurement-based approaches, while realistic, struggle to generalize across different workloads and peripheral configurations. Static analyses offer stronger guarantees but demand accurate models of each system component.

The stateful nature of peripherals is a particular obstacle for static analysis. Existing static analyses are insufficient in two key ways: (1) peripherals are either unmodeled or modeled under overly restrictive assumptions, limiting applicability to a narrow class of devices and programs; and (2) existing tools impose significant workflow burden, requiring external tool integration or custom peripheral implementations from the programmer such as \cite{woca}.

To address these limitations, we use the notion of \textit{peripheral typestate}, a type-level representation of a peripheral's modes of operation, and combine it with prior work on automatic cost analysis in type systems. The result is, to our knowledge, the first type system to unify these two ideas, enabling symbolic energy bounds for embedded systems without invasive changes to programmer workflow. By embedding peripheral typestate directly into the type system, we model a broad class of peripherals without restrictive assumptions, and since the analysis is part of the type system, programmers need not integrate external tools or provide custom peripheral implementations to obtain energy bounds.
% \milijana{Given the happy accident of suddenly having more space, you could probably add another sentence in this paragraph to say more about how this approach will overcome the challenges}

\section{Background and Related Work}
We organize related work around three pillars: WCEC analysis, peripheral typestate, and resource-aware type systems, as well as an approach that combines some of these ideas. 

Our goal is to provide a tight, peripheral-aware bound on the WCEC of embedded systems. Recent prior work for this problem has typically followed the same high level structure: enumerate all possible runtime behaviors, and then translate to a formal constraint-based problem that can be solved with knowledge about the specific hardware that is being executed. \cite{Wagemann2022, raffeck_et_al_OASIcs_WCET_2019_4,wegener_et_al_OASIcs_WCET_2023_9, 8847394} Typically, this enumeration is done via a static analysis. An approach that stands out is \textit{WoCA} \cite{woca} which enhances this enumeration by integrating peripherals and their modes of operation. In order for the analysis to be sound and provide accurate bounds, code must be written in a way that only activates one peripheral device at a time. Additionally, \textit{WoCA} requires custom device implementations for using their analysis, and does not consider processor energy costs. This means that programs that rely on the processor and do not use peripherals extensively will have incorrect bounds. 

Another approach for analyzing resource consumption is through the usage of type systems. An established technique used is known as Automatic Amortized Resource Analysis (AARA) \cite{Hoffmann_Jost_2022}. The analysis uses type inference rules whose derivations serve as proofs of resource bounds. It first performs standard type inference and then infers resource annotations by solving a linear program based on the potential method of amortized analysis, with soundness proven via a cost semantics that defines program resource consumption. This has been used to obtain bounds on the consumption of a variety of resources such as time and memory usage for free, motivating our idea of building a type system for energy analysis. Such a type system can overcome the challenge of integrating a static analysis, while also providing robust WCEC bounds. However, AARA alone is insufficient. The authors of \cite{Hoffmann_Jost_2022} note that a major weakness of using this approach is that it lacks knowledge of peripheral behavior, which leaves it unable to address the first challenge of many prior WCEC works.

% \milijana{It doesn't motivate the *notion* of peripheral typestate, but rather our *use* of typestate. typestate as a notion 
% has existed for a long time. You can rework this flow to say there's an existing notion to describe the peripheral behaviour that AARA doesn't account for, typestate. More specifically, typestate is [xyz]. then describe how abacus fits in. }
This motivates our use of \textit{peripheral typestate}. AARA and similar approaches lack knowledge of the internal state of peripherals, which is highly complex and directly affects their behavior and energy consumption. An existing notion for capturing exactly this kind of behavior is \textit{typestate}: a type-level representation of the internal state of a stateful object. The Abacus framework \cite{abacus} applies this notion directly to peripherals, providing a Rust-embedded DSL that lets programmers specify peripheral states and valid operations, then automatically generates peripheral typestates enforced by the compiler. By incorporating the idea of typestate into our type system, we can address the modeling gap that AARA alone leaves open.

Our goal is to derive tight symbolic energy bounds for embedded systems by tracking peripheral state directly within a type system in the style of AARA, obtaining these bounds for free. To our knowledge, the closest prior work to this idea is Performal \cite{performal}, which analyzes latency in distributed systems by modeling nodes as state machines. Like peripherals, distributed nodes have complex internal state whose behavior and transition costs depend heavily on current state, and Performal derives latency bounds by tracking both. However, rather than adopting an external proof framework like what Performal does, we integrate this analysis directly into a type system.

\section{Approach and Uniqueness}
\subsection{Approach Overview}
Our type system has two main goals: keeping track of peripheral state as accurately as possible, and using this information to provide symbolic energy cost bounds. Our type system achieves these goals through a novel approach to using peripheral typestate in our cost analysis. 

To derive precise symbolic energy bounds, the analysis must account for several sources of energy consumption. These sources can be split into three main categories: costs from executing statements of code on the processor, costs from using peripherals, and costs from active peripherals in the background. The usage of typestate helps us to reason about the second and third cost categories in particular, as peripherals use a different amount of power based on the current mode of operation. A naive approach that assumes worst-case power consumption throughout would produce bounds that are sound but far too conservative to be useful. Precise symbolic bounds therefore require knowing exactly which state each peripheral occupies at each program point.

To track peripheral state precisely, we start by noticing that many peripheral states obey a natural implicit ordering. For example, in order for a sensor to be initialized, it must be turned on. To encode this, we represent the typestate of a peripheral as a lattice of states, where $s_1 \leq s_2$ means that a peripheral in state $s_1$ is allowed to transition to state $s_2$. We represent this in our code by \texttt{transition}$(p, st)$. However peripherals do not solely move up the lattice. Commonly, peripherals return to a prior state upon completing an operation. We model these ``backwards transitions'' via downgrade policies, $s_2 \hookrightarrow s_1$. These are represented in our code by \texttt{downgrade}$(p, st)$. 

While the lattice gives us the structure to reason about peripheral states, requiring explicit annotations at every transition would be burdensome and error-prone. Instead, we use type inference. Each time a peripheral transitions to a new state, we introduce a fresh state variable $\alpha$---analogous to type variables---and push it onto a stack associated with that peripheral. Additionally, we  add constraints to ensure that the state variable represents a valid peripheral state for the transition. We refer to the process of substituting state variables for concrete states as ``concretization.'' When we are done analyzing a function, we concretize all state variables except the one at the bottom of the stack, which we refer to as eager concretization. The bottom-of-stack variable represents the peripheral's state at function entry. Therefore, it is concretized at the call site, where the caller has precise knowledge of the peripheral's initial state. We refer to this as lazy concretization. This prevents overly conservative bounds: eagerly resolving intermediate states avoids unnecessary uncertainty, while deferring entry-state concretization keeps function costs parametric until the call-site, where more precise information may be available. We illustrate this using a running example.

\paragraph{Running Example.}
The program in the left column of Figure~\ref{fig:example} models a simple task using a temperature sensor as a peripheral. The lattice for the temperature sensor is as follows: $\langle \texttt{Off}, \texttt{Idle}, \texttt{Measure} \rangle$ where $\texttt{Off} \leq \texttt{Idle} \leq \texttt{Measure}$. It also has downgrade policies $\texttt{Measure} \hookrightarrow \texttt{Off}$ and $\texttt{Idle} \hookrightarrow \texttt{Off}$. The function \texttt{log\_data} takes a temperature sensor, and then transitions to the \texttt{Measure} state. Afterwards, it performs a measurement and conditionally writes to a variable. Finally, it powers the sensor down before returning. In the main function, we initialize the temperature sensor, and then call the \texttt{log\_data} function. We use this example to demonstrate how our type system tracks peripheral state and derives symbolic energy bounds using the right column. 
\begin{figure}
\centering
\small
\setlength{\tabcolsep}{3pt}
\begin{tabular}{|r|p{4cm}|l|}
\hline
\textbf{} & \textbf{Code} & \textbf{Cost Expression} \\
\hline
1 & \texttt{fn log\_data(temp) -> u64 \{} & \\
2 & \ \texttt{transition(temp, Measure)} & $C(\text{temp}: (\alpha_0, \text{Measure}))$ \\
3 & \ \texttt{let res = temp.measure()} & $C(\text{temp}: \text{Measure})$ \\
4 & \ \texttt{let t = 0} & $C_v + C(\text{temp}: \text{Measure})$ \\
5 & \ \texttt{if(res == 65) \{} & \\
6 & \quad \texttt{let t = 1} & $\text{max}(C_v, 0) + C(\text{temp}: \text{Measure})$ \\
7 & \ \texttt{\}} & \\
8 & \ \texttt{downgrade(temp, Off)} & $C(\text{temp}: (\text{Measure},\text{Off}))$ \\
9 & \ \texttt{return t} & \\
10 & \texttt{\}} & \\
\hline
\multicolumn{3}{|l|}{$\texttt{log\_data} \text{ cost} = C_l = C(\text{temp}: (\alpha_0, \text{Measure})) + 3C(\text{temp}: \text{Measure})$} \\
\multicolumn{3}{|l|}{$\phantom{C_l} + \ C_v + \text{max}(C_v, 0) + C(\text{temp}: (\text{Measure}, \text{Off}))$} \\
\hline
1 & \texttt{fn main() \{} & \\
2 & \ \texttt{transition(temp, Idle)} & $C(\text{temp}: (\alpha_1, \text{Idle}))$ \\
3 & \ \texttt{let t = log\_data(temp)} & $C_l[\alpha_0 \mapsto \text{Idle}]$ \\
4 & \texttt{\}} & \\
\hline
\multicolumn{3}{|l|}{\textit{Total cost} $= C(\text{temp}: (\alpha_1, \text{Idle}))[\alpha_1 \mapsto \text{Off}]$} \\
\multicolumn{3}{|l|}{$\phantom{\text{Total cost}} + \ C(\text{temp}: (\text{Idle}, \text{Measure})) + 3C(\text{temp}: \text{Measure})$} \\
\multicolumn{3}{|l|}{$\phantom{\text{Total cost}} + \ C_v + \text{max}(C_v, 0)$} \\
\multicolumn{3}{|l|}{$\phantom{\text{Total cost}} + \ C(\text{temp}: (\text{Measure}, \text{Off}))$} \\
\hline
\end{tabular}
\caption{Running Example with Cost Analysis}
\label{fig:example}
\end{figure}

As seen in the right-hand column of Figure~\ref{fig:example}, we express energy bounds as symbolic cost expressions parameterized by peripheral state. Function costs remain parametric in the input state $\alpha$ until lazy concretization occurs. We generally have 3 types of cost expressions. $C(p: st)$, represents the cost of a peripheral being in a concrete state $st$, with $C(p: \alpha)$ being the unknown state version. $C(p: (st, st'))$ represents the cost of transitioning from one state to another, with $C(p: (\alpha, st'))$ being the unknown version. Our language always requires an explicit state for the target of a transition, so we always know at least the intended destination state. Finally, $C[\alpha \mapsto st]$ represents concretizing $C$ with state $st$. The full formal grammar is given in Figure~\ref{fig: symolic_energy_syntax}. 
\vspace{-10px}
\begin{figure}[h]
    \centering
   \[
    \begin{array}{lcll}
     \text{Cost expression} & C  \bnfdef & C(p : st) \\ & \bnfalt & C(p : (st, st'))  \\ & \bnfalt & C(p : (\alpha, st))  \\
     & \bnfalt & C(p : \alpha)  \\ & \bnfalt & \m{max}(C_1, C_2) \ \\ & \bnfalt & C_1 + C_2   \\ & \bnfalt & C[\alpha \mapsto st] 
\end{array}
   \]
   \caption{Symbolic Energy Cost Syntax}
   \label{fig: symolic_energy_syntax}
\end{figure}
\subsection{Type System.} 

We present our type system on a simple imperative language in Figure~\ref{fig: syntax} which includes explicit \texttt{transition} and \texttt{downgrade} statements for performing the state transitions as discussed previously. 

\begin{figure}[h]
    \[
\begin{array}{lcll}
\text{Peripheral Declarations: }p \bnfdef \ \text{id} \{\langle s_{0},s_{1},\ldots,s_{n}\rangle, <, st, D \}, p \ \vert \ \cdot 
\end{array}
\]
\[
\begin{array}{lcll}
\text{Declarations: }d & \bnfdef & \text{fn } f(x, p): \tau_1 \rightarrow \tau_2 \ \{c; ret := e\}, d \ \vert \ \cdot
\end{array}
\]
\[
\begin{array}{lcll}
\text{Commands: }c & \bnfdef & \text{let } x = e & \\ & \bnfalt & \text{let } x = f(e, p) \\ & \bnfalt & \text{if } e \text{ then } c_1 \text{ else } c_2  \\ & \bnfalt & \text{transition}(x, \text{st})  \\ & \bnfalt & \text{downgrade}(x, \text{st}) \\ & \bnfalt & \text{clock}()  \\ & \bnfalt & \text{match }p \text{ with } \text{st}_1 \rightarrow c_1 | \dots | \text{ st}_n \rightarrow c_n  \\ & \bnfalt & c_1; c_2 
\end{array}
\]
\[
\begin{array}{lcll}
\text{Expressions: }e & \bnfdef & x \\
    & \bnfalt & v  \\
    & \bnfalt & e_1 \text{ binop } e_2 
\end{array}
\]
\caption{Syntax}
\label{fig: syntax}
\end{figure}

Our type system uses the following additional constructs: 

\[
\begin{array}{llcl}
\textit{Typing Context} & \tctx \bnfdef \ x \rightarrow \tau, \tctx \bnfalt \cdot
\\ 
\textit{Types} & \tau \bnfdef \ \text{int} \bnfalt \text{bool} \bnfalt \text{unit} \bnfalt \rho \bnfalt \tau_1 \rightarrow \tau_2 
\\
    \textit{State Constraints} & \varphi \bnfdef \ \alpha = \Vec{st} \bnfalt \alpha \leq st \bnfalt \alpha \subseteq \Vec{st} \bnfalt \\ & \varphi_1 \wedge \varphi_2 \bnfalt \varphi_1 \vee \varphi_2 \\
     \textit{Constraints} & \Phi  \bnfdef \ \varphi, \Phi \bnfalt \cdot  \\
\textit{Peripheral Context} & \pctx \bnfdef \  p \rightarrow [\alpha], \pctx \bnfalt \cdot
\end{array}
\]

We map each peripheral to a \textit{stack} of these state variables to support the lazy/eager concretization strategy mentioned before. Our typing judgment takes the following form: $\tctx, \pctx, \Phi, C \vdash c \rightarrow \tctx', \pctx', \Phi', C'$. The judgment means that given a typing context, peripheral context, a set of current constraints and a current cost, evaluating $c$ can update all four of these constructs. 
We present selected typing rules below:
\begin{mathpar}
\small
\inferrule*[left=T-Seq]
{
\tctx,\pctx,\Phi, C \vdash c_1 :
\tau_1
\rightarrow
\tctx',\pctx', \Phi', C' \\ \tctx',\pctx',\Phi', C' \vdash c_2:
\tau_2
\rightarrow
\tctx'',\pctx'', \Phi'', C'' \\ C^* = C + \m{cost}(\pctx) 
}
{
\tctx,\pctx, \Phi, C \vdash
c_1 ; c_2 :
\tau_2
\rightarrow
\tctx'',\pctx'', \Phi'', C^*
}

\inferrule*[left=T-Transition] 
{ \tctx, \pctx, \Phi, C \vdash x : \rho \\ \pctx(x) = \alpha_0 \\ \alpha_1 \text{ fresh} \\ \Phi' = \Phi \land \alpha_0 \leq st \land \alpha_1 = st \\ \pctx'[x] = \alpha_1 :: \pctx[x] \\ C' = C + C(p : (\alpha_0, st))
} 
{ 
\tctx, \pctx,\Phi, C \vdash \text{transition}(x, st) : \text{unit} \rightarrow \tctx, \pctx', \Phi', C' 
}

\inferrule*[left=T-Downgrade]
{
\tctx, \pctx, \Phi, C \vdash x : \rho \\ s \hookrightarrow st \in x(D) \\ \pctx(x) = \alpha_0 \\ \alpha_1 \text{ fresh} \\ \Phi' = \Phi \land \alpha_0 = s \land \alpha_1 = st \\ \pctx'[x] = \alpha_1 :: \pctx[x] \\ C' = C + C(p : (\alpha_0, st)) 
}
{
\tctx,\pctx, \Phi, C \vdash
\text{downgrade}(x,st) :
\text{unit}
\rightarrow
\tctx, \pctx', \Phi',C'
}

\inferrule*[left=T-FunDecl] 
{ \alpha_{\text{in}} \text{ fresh} \\ \pctx = p : [\alpha_{\text{i}}] \\ \tctx, \pctx, \cdot, \cdot \vdash c \rightarrow \pctx_f, \Phi_f, C_f \\ \pctx_f(p) = \alpha_{\text{o}} \\ \Vec{st} = \m{solve}(\Phi_f \setminus \Phi_f|_{\alpha_i}) \\ C = C_f[\Vec{st}] \\ \Vec{s_1} = \m{solve}(\Phi_f|_{\alpha_{\text{i}}}) \\ \Vec{s_2} = \m{solve}(\Phi_f|_{\alpha_{\text{o}}}) \\ \tau_f = \tau_1 \rightarrow \tau_2
} 
{ \tctx \vdash \text{fn } f(x, p) \ \{c; ret := e\} : \tau_f, C, \Phi_f|_{\alpha_i}, \Vec{s_i}, \Vec{s_o} }

\inferrule*[ left=T-Call]
{
\tctx(f)=
\tau_1 \rightarrow \tau_2, C_f, \Phi_f|_{\alpha_i}, \Vec{s_i}, \Vec{s_o}
\\ \alpha = \pctx(p) \\ \m{rename}(\Phi_f|_{\alpha_i}, \alpha) \\ 
\tctx,\pctx,\Phi, P \vdash e : \tau_1, C_a \\ \alpha' \text{ fresh} \\ \Vec{s} = \m{solve}(\Phi_f|_{\alpha_i} \land \Phi|_{\alpha}) \\ \Vec{s} \subseteq \Vec{s_1} \\ \pctx'[p] = \alpha' :: \pctx[p] \\ C' = C_f[\Vec{s}] + C_a + C \\ \Phi' = \Phi \land \alpha \subseteq \Vec{s_1} \land \alpha' \subseteq \Vec{s_2} 
}
{
\tctx,\pctx, \Phi, C \vdash
\text{let } x = f(e,p) :
\text{unit}
\rightarrow
\tctx[x:\tau_2], \pctx', \Phi', C'
}

\end{mathpar}
\begin{mathpar}
    \small

\end{mathpar} 

\textsc{T-Seq} is our primary rule for composing costs. While we can simply thread the cost through, as mentioned before we need to consider the cost of active but non-operating peripherals. We represent this background cost by $\m{cost}(\pctx)$, which is a shorthand for the summation of the current state variable of each active peripheral. In our example, since we only have the temperature sensor, the background cost is represented by the addition of $C(\text{temp}: \texttt{Measure})$. 

\textsc{T-Transition} and \textsc{T-Downgrade} are  rules for peripheral state updates. Both rules maintain an SSA-like property, creating a fresh state variable and adding corresponding constraints on each peripheral state update. When we add the cost to our energy expression, we add a cost expression parametric in the current state variable, to be concretized later. In our running example, this is represented by $C(\text{temp}: (\alpha_0, \texttt{Measure}))$

\textsc{T-FunDecl} is the most complex rule in the type system. We want to derive a symbolic energy expression describing the behavior of the function. We eagerly concretize state variables that do not depend on the initial state, since their behavior is fully known within the declaration itself. Upon completing the analysis, we return four additional pieces of information alongside the function type: the expected input and output peripheral states, the unsolved constraints, and the partially concretized energy expression. The unsolved constraints are carried forward so that cost expressions can be fully concretized at call sites.

\textsc{T-Call} completes the analysis started in \textsc{T-FunDecl}. We rename the constraints carried from the function declaration to refer to the current peripheral state variable, then combine them with its current constraints. In our running example, this renaming allows us to use the knowledge that \texttt{temp} is \texttt{Idle}, yielding a fully concretized cost expression $C(\text{temp: (Idle, Measure)})$ for the initial transition. 

For any cost expression that has not been concretized after analyzing all the functions, we concretize it to the highest state that respects all of its  constraints to obtain the tightest bound. In the running example, the very first transition in \texttt{main} will not be concretized since it is not called anywhere, so we set $\alpha_1 = \texttt{Off}$.

\section{Results and Contributions}
Combining everything together gives a type system that automatically derives symbolic cost expressions for embedded systems with stateful peripherals, requiring no programmer annotations beyond peripheral lattice definitions. While we are effectively able to model stateful devices and processors, a significant source of processor energy cost depends on the clock frequency. Peripheral operations can change the clock speed, affecting processor costs. Our primary next steps are to enhance our analysis to account for varying clock speeds imposed by these operations. We include $\m{clock}()$ as a language construct to signal to the type system that the processor clock speed changes at this point. We can use this to separate programs into regions which are determined by a static clock speed. Then we can build up cost expressions for each region, and combine them together to create a \textit{symbolic energy property} that combines cost expressions across clock speed regions, capturing the total energy cost with varying clock speed. Also, we are currently implementing this type system within the Abacus framework using Rust's proc-macros system \cite{rust_procedural_macros_reference}.We aim to integrate with standard peripherals and the Tock Real Time Operating System \cite{Levy2017-ii} to test the viability of our approach on real embedded systems. 

\newpage
\bibliographystyle{plain}
\bibliography{ref}

 
\end{document}
